---
title: "Pset 1 - Water usage"
author: "425/625"
date: "Spring 2024"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(pubtheme)
```

## Introduction 

Water scarcity is a major issue in many parts of the world. According to the [United Nations](https://www.un.org/en/climatechange/science/climate-issues/water), "About two billion people worldwide don't have access to safe drinking water today ([SDG Report 2022](https://unstats.un.org/sdgs/report/2022/Goal-06/?_gl=1*aquqlc*_ga*MTAzMzA2ODI0Mi4xNzA0NjYzNjY3*_ga_TK9BQL5X7Z*MTcwNDY2NzIyMS4yLjEuMTcwNDY2NzI1My4wLjAuMA..)), and roughly half of the world’s population is experiencing severe water scarcity for at least part of the year ([IPCC](https://www.ipcc.ch/report/ar6/wg2/downloads/outreach/IPCC_AR6_WGII_FactSheet_FoodAndWater.pdf)). These numbers are expected to increase, exacerbated by climate change and population growth ([WMO](https://public.wmo.int/en/media/press-release/protect-our-people-and-future-generations-water-and-climate-leaders-call-urgent))."


In this problem set, we will investigate water usage estimates by crop in the United States. The `.csv` for this data set comes from [here](https://databank.illinois.edu/datasets/IDB-4607538) (by checking Select All and clicking Get Custom Zip) and the associated academic journal article is [here](https://agupubs.onlinelibrary.wiley.com/doi/10.1029/2022WR032804). See [this thread](https://twitter.com/MeganKonar/status/1618943789279875074?s=20&t=nrjRJQ2tuEO0GyTwCK-XVA) on X for a summary.  

Read the academic article to familiarize yourself with the basics of the water usage data. You don't need to know how these water usage levels were estimated, so you can skip over those parts. We are going to focus on visualizing the water levels using the estimates that they generated. 

## Data preparation

The `.zip` file `rawdata/DOI-10-13012-b2idb-4607538_v1.zip` contains one `.csv` file per source (SWW, GWW, GWD) per year from 2008 to 2020. There are also a couple of `.txt` files in the folder. We can use `unzip` with `list = TRUE` to see what's in the `.zip` file. 

```{r}
unzip(zipfile = 'rawdata/DOI-10-13012-b2idb-4607538_v1.zip', 
      list = TRUE) ## this lists the filename, but does not unzip the file
```

Before summarizing/visualizing this data, we'll want to join these data sets. We could certainly unzip the file manually. We can also do this in R using `unzip`. 

```{r eval=F}
unzip(zipfile = 'rawdata/DOI-10-13012-b2idb-4607538_v1.zip', 
      junkpaths = TRUE, 
      exdir = 'rawdata') ## gets rid of paths, keeps only filenames 
```


#### 1. Join data
First, let's create a data set with all years/crops together in one data frame. Below is some code to help you get started. Add comments to each place there is `##` to explain what the chunk of code is doing. Then add code to the `Tranforming data` Section to transform the data into a data frame with 5 columns: `GEOID`, `crop`, `source`, `year`, and `value` (indicating km^3 of water).

Note that `eval = F` at the start of the chunk will prevent this chunk from evaluating when you knit the document. You can temporarily remove it if you'd like, but you'll want to add it back before knitting the document so that knitting takes less time.

```{r eval = F}
# SWW: surface water withdrawals
# GWW: groundwater withdrawals
# GWD: nonrenewable groundwater depletion
# GWA: groundwater abstractions
sources = c('gwd', 'sw', 'gwa')
years = 2008:2020
d = NULL

for(s in sources){
  cat(s, '') ## show progress
  
  for(year in years){
    cat(year, '') ## show progress
  
    ## read each file
      # filename: each file begins with 'rawdata..._v1/'
        # s is either gwd, sw, or gwa and year ranges from 2008 to 2020
        # (specific s and year values are specified by the for loops)
      # each file name is then read into df
        # so we have one data frame with data from all sources and years
    filename = paste0('rawdata/DOI-10-13012-b2idb-4607538_v1/DOI-10-13012-b2idb-4607538_v1/', s, '_', year, '.csv')
    df = read.csv(filename)
    head(df)
    
    
    ## Tranform data #######################################
    ## Use `pivot_longer`, `separate`, and/or other functions to transform this
    ## data frame into a data frame with 5 columns: 
    ## GEOID, crop, source, year, and value (indicating km^3 of water)
    
    df = df %>% 
      pivot_longer(cols = !GEOID, 
                   names_to = 'src.crop.year', 
                   values_to = 'value') %>% 
      separate(src.crop.year, into = c('src', 'crop', 'year'), sep = '[.]') %>%
      relocate(GEOID, crop, src, year, value)
    df
    
    
    
    ## end of transforming data ############################
    
    ## Each df has values for only one source and year
      # so at the end of each year iteration, we add df to the larger d dataframe
      # after all iterations of source and year, d contains information for all sources and all years
      # (this is why each df has 64460 observations and d has >2.4 million)
    d = rbind(d, df)
  }
  
  cat('\n') ## start a new line before showing progress for the next source
}
head(df)
tail(d)


```



## Data exploration and summaries

Let's load the data we'll use for the rest of the assignment. This is the data set created in #1, so if you were unable to finish #1, you can still do the rest of the assignment. 

```{r}
d = readRDS('data/water.usage.rds')
head(d)
```


#### 2. Summaries of data 

Find the mean, the change from 2008 to 2020, and the percent change from 2008 to 2020, for each crop and each source (SWW, GWW, GWD). 

```{r}

# value = km^3 of water
# mean for each crop and each source = sum of value for all crops / number of crops
  # (for each crop, src)
  
# total water usage for each crop, source, year, for all census tracts
dd3 = d %>% 
  group_by(crop, src, year) %>% 
  summarise(value = sum(value))
dd3

# avg annual water usage for each crop and source
dd4 = dd3 %>% 
  group_by(crop, src) %>% 
  mutate(mean = mean(value)) %>% 
  distinct(mean)
dd4

# change
dd5 = dd3 %>% 
  group_by(crop, src) %>% 
  mutate(change = last(value) - first(value)) %>% 
  distinct(change)
dd5

# percent change
dd6 = dd3 %>% 
  group_by(crop,src) %>% 
  mutate(percent = (last(value) - first(value)) / first(value)) %>% 
  distinct(percent)
dd6

```



## 3. Convert Table 2 to a visualization

Create a visual representation of the information in Table 2. Create a visualization (or visualizations) that contains mean, change, and percent change in water usage from each crop and source. 

```{r}
# combine mean, change, percent change into 1 dataframe
table2 <- dd3 %>% 
  group_by(crop, src) %>% 
  mutate(mean = mean(value),
         change = last(value) - first(value),
         percent = (last(value) - first(value)) / first(value)) %>% 
  distinct(mean, change, percent)

table2

```

```{r}

# contains mean, change, percent change
# from each crop and source

# Mean
meanvis <- table2 %>% ggplot(aes(y = crop,
                  x = mean,
                  fill = src)) +
  geom_col(width = .8, position = position_dodge()) +
  theme_pub() +
  theme(text = element_text(size = 5),
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 8),
        legend.position = "right") +
  labs(title = "Average Water Usage by Crop and Source",
       x = "Average Water Usage (km^3)",
       y = "Crop",
       fill = "Source") +
   scale_fill_discrete(labels = c("GWA", "GWD", "SW")) +
  facet_grid(src~.)

# Change
changevis <- table2 %>% ggplot(aes(y = crop,
                  x = change,
                  fill = src)) +
  geom_col(width = .8, position = position_dodge()) +
  theme_pub() +
  theme(text = element_text(size = 5),
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 8),
        legend.position = "right") +
  labs(title = "Change in Water Usage from 2008 to 2020 by Crop and Source",
       x = "Change in Water Usage from 2008 to 2020 (km^3)",
       y = "Crop",
       fill = "Source") +
   scale_fill_discrete(labels = c("GWA", "GWD", "SW")) +
  facet_grid(src~.)

# Percent Change
percentvis <- table2 %>% ggplot(aes(y = crop,
                  x = percent,
                  fill = src)) +
  geom_col(width = .8, position = position_dodge()) +
  theme_pub() +
  theme(text = element_text(size = 5),
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 8),
        legend.position = "right") +
  labs(title = "Percent Change in Water Usage",
       x = "Percent Change in Water Usage",
       y = "Crop",
       fill = "Source") +
   scale_fill_discrete(labels = c("GWA", "GWD", "SW")) +
  facet_grid(src~.)

meanvis
changevis
percentvis

```


# Figure 4

Figure 4 shows the average water usage by crop and source.

- A. average irrigation water usage by source, colored by crop, 
- B. average irrigation water usage by crop, colored by source 

Two other options for visualizing a numeric variable broken down by two different categorical variable would be a tile plot/grid plot (e.g. https://github.com/bmacGTPM/pubtheme?tab=readme-ov-file#grid-plot) and a mosiac plot (https://haleyjeppson.github.io/ggmosaic/).


## 4. Create a tile plot/grid plot of the data in Figure 4. 

```{r}

g <- table2 %>% ggplot(aes(x = src,
                      y = crop,
                      fill = mean)) +
  geom_tile(linewidth = 0.4,
            show.legend = T,
            color = pubdarkgray) +
  labs(title = "Average Water Usage by Crop and Source",
       x = "Source",
       y = "Crop",
       fill = 'Mean') 

g %>% pub(type = 'grid',
          xbreaks = seq(2, 32, by=2)) 

```

## 5. Create a mosiac plot of the data in Figure 4. 

```{r}

library(ggmosaic)

table2 %>%
  ggplot() +
  geom_mosaic(aes(x = product(src), fill = crop, weight = mean), show.legend = F) + 
  theme_mosaic() + 
  theme(text = element_text(size = 5),
        axis.text.y = element_text(angle = 45, hjust = 1),
        axis.title = element_text(size = 8),
        title = element_text(size = 10)) +
  labs(title = "Average Water Usage by Crop and Source",
       x = "Source",
       y = "Crop") 

```


## 6. What are the benefits (other than it fits on one plot) and drawbacks of these two plots?

Tile plot: It is easy to see the crops and sources where water usage is most heavily concentrated, but the data is concentrated strongly in one particular tile. This makes it more difficult to compare the data across other tiles.

Mosaic plot: The size variation makes it easy to compare water usage across crops and sources. It is difficult, however, to understand water usage for crops that don't use much water, because the mosaic tiles are too small to see and compare them. 

## 7. Figure 6 

Figure 6 uses a different color scale for each plot. Discuss the benefits and drawbacks of this choice. What was the main purposes of this figure? Given the main purpose, would you recommend using the same color scale, or different color scales, for each plot?

Using different color scales makes it easier to see the location and concentration of water use for each crop and source combination. If every color scale ranged from the minimum to maximum value (0.001 to 1.6), it would be more difficult to see the variation in graphs with data concentrated at one end of the spectrum, such as the soy-gwd and other animal feed-gwd plots. But the variation in scale also makes it more difficult to compare data across plots - although cotton and other produce appear to have the same concentration of irrigation from groundwater abstractions, other produce actually uses significantly more water. 

The purpose of the figure is to show the location and concentration of irrigation for each source and crop. The figure is not meant to be used to compare one crop or water source to another; it is meant to provide information about each individual crop-source combination. Given this purpose, I would recommend using the same color scale for each plot. 


## 8. Figure 8 

Figure 8 also uses a different color scale for each plot.  Discuss the benefits and drawbacks of this choice. What was the main purposes of this figure? Given the main purpose, would you recommend using the same color scale, or different color scales, for each plot?

As in Figure 6, using different color scales makes it easier to see the variations in data for each plot, but it also makes it difficult to compare data between plots. 

The figure's purpose is to show the difference between estimated and reported withdrawals and compare that difference between surface water and groundwater and between 2010 and 2015. Since the figure is meant to be used for comparison, I would recommend using the same color scale for each plot. 


## 9. Breakdown of GWW
The paper notes in Section 3.1 that $GWW = GWW_{sustainable} + GWW_{unsustainable}$, and that $GWD = GWW_{unsustainable}$. Create a visualization showing the percent of GWW that is GWD for each crop. Use the mean values for water usage. 


```{r}

t3 <- table2 %>% 
  filter(src == "gwa" | src == "gwd") %>% 
  group_by(crop) %>% 
  mutate(gww = sum(mean)) %>% 
  filter(src == "gwd") %>% 
    mutate(gwdpercent = mean / gww * 100)

t3 %>% ggplot(aes(x = crop, 
           y = gwdpercent,
           fill = src)) +
  geom_col() +
  theme_pub() +
  theme(axis.text.x = element_text(angle = 45, 
                                   hjust = 1),
        legend.position = "none") +
  labs(title = "GWD as Percentage of GWW", 
       x = "Crop",
       y = "Percentage")

```


## 10. Custom visualization
What is another question you have about this data? Create a visualization that attempt to answer your question. 

For each crop, what is the ratio of average water usage for each water source compared to the total average  water usage?

```{r}

table2 %>% ggplot(aes(x = mean,
                      y = crop,
                      fill = src)) +
  geom_col() +
  theme_pub() +
  labs(title = "Average Water Usage for Each Source as 
       Fraction of Total Average Water Usage",
       x = "Mean (km^3)", 
       y = "Crop",
       fill = "Source") +
  theme(axis.text.y = element_text(size = 9),
        legend.position = "right") +
  scale_fill_discrete(labels = c("GWA", "GWD", "SW"))

```

